---
title: "Project Proposal data 606 for NBA teams"
author: "Leon Lin"
date: "4/11"
output:
  pdf_document: default
  html_document: default
---



## Research Question

**Do NBA team advanced metrics like Net Rating, True Shooting Percentage (TS%), and Pace significantly predict regular season win percentage?**

The goal is to determine which advanced statistics are most strongly associated with a higher winning percentgae in the NBA. 

We will use a multiple linear regression model using these predictors to see whetehr we can reliably estimate team win percentage in a given season.

---

## Cases

- **Unit of observation:** One NBA team in one regular season 
- **Estimated number of cases:** around 750 observations 
- Seasons with fewer than 30 teams will have slightly fewer cases per year

---

## Data Collection Method

- Data's from **Basketball Reference** (basketball-reference.com) the official NBA box scores and game logs
- Advanced statistics are derived metrics calculated from raw box score data using standard NBA and statistical formulas
- The dataset can be downloaded as a CSV from Basketball Reference's Team Advanced Stats table:  
  [https://www.basketball-reference.com/leagues/NBA_stats_per_game.html](https://www.basketball-reference.com/leagues/NBA_stats_per_game.html)
- Data can also be pulled directly in R using the package hoopr

---

## Type of Study

- **Observational study** data reflect actual team performance across different NBA regular seasons

---

## Data Source

Basketball Reference. (2024). *NBA & ABA Team Advanced Stats*. Sports Reference LLC. Retrieved from [https://www.basketball-reference.com/leagues/](https://www.basketball-reference.com/leagues/)

---

## Response Variable
variable `win_pct`
Description Regular season win percentage
Type numerical/continous 
---

## Explanatory Variables

| Variable | Description | Type |
|---|---|---|
| `net_rtg` | Net Rating is points scored minus points allowed per 100 possessions | **Numerical (continuous)** |
| `ts_pct` | True Shooting %: shooting efficiency takes into accounting for 2pointers, 3points, and free throws | **Numerical (continuous)** |
| `pace` | Pace is estimated number of possessions per 48 minutes | **Numerical (continuous)** |


---

## Relevant Summary Statistics

```{r load-data}
library(hoopR)
library(dplyr)
library(ggplot2)


set.seed(42)
n <- 750

nba_demo <- data.frame(
  season    = rep(2000:2024, each = 30),
  team      = rep(paste0("Team_", 1:30), times = 25),
  win_pct   = pmin(pmax(rnorm(n, mean = 0.500, sd = 0.120), 0.1), 0.9),
  net_rtg   = rnorm(n, mean = 0.0, sd = 5.5),
  ts_pct    = rnorm(n, mean = 0.545, sd = 0.025),
  pace      = rnorm(n, mean = 97.5, sd = 3.5)
)

nba_demo$win_pct <- 0.500 + 0.035 * nba_demo$net_rtg +
                    0.8   * (nba_demo$ts_pct - 0.545) +
                    rnorm(n, 0, 0.04)
nba_demo$win_pct <- pmin(pmax(nba_demo$win_pct, 0.10), 0.90)
```

```{r summary-stats}

nba_demo %>%
  select(win_pct, net_rtg, ts_pct, pace) %>%
  summary()
```

```{r summary-table}
library(knitr)

sum_table <- nba_demo %>%
  select(win_pct, net_rtg, ts_pct, pace) %>%
  summarise(across(everything(),
    list(Mean = mean, SD = sd, Min = min, Max = max),
    .names = "{.col}_{.fn}"
  ))


data.frame(
  Variable  = c("Win %", "Net Rating", "True Shooting %", "Pace"),
  Mean      = c(mean(nba_demo$win_pct), mean(nba_demo$net_rtg),
                mean(nba_demo$ts_pct), mean(nba_demo$pace)),
  SD        = c(sd(nba_demo$win_pct), sd(nba_demo$net_rtg),
                sd(nba_demo$ts_pct), sd(nba_demo$pace)),
  Min       = c(min(nba_demo$win_pct), min(nba_demo$net_rtg),
                min(nba_demo$ts_pct), min(nba_demo$pace)),
  Max       = c(max(nba_demo$win_pct), max(nba_demo$net_rtg),
                max(nba_demo$ts_pct), max(nba_demo$pace))
) %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  kable(caption = "Summary Statistics for Key Variables (Demo Data)")
```

```{r plot-net-rtg}
ggplot(nba_demo, aes(x = net_rtg, y = win_pct)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Net Rating vs. Win Percentage (NBA Team-Seasons)",
    subtitle = "Each point = one team-season",
    x = "Net Rating (pts per 100 possessions)",
    y = "Win Percentage"
  ) +
  theme_minimal()
```

```{r plot-ts}
ggplot(nba_demo, aes(x = ts_pct, y = win_pct)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "True Shooting % vs. Win Percentage (NBA Team-Seasons)",
    x = "True Shooting %",
    y = "Win Percentage"
  ) +
  theme_minimal()
```

```{r correlations}
# Correlation matrix
cor(nba_demo[, c("win_pct", "net_rtg", "ts_pct", "pace")]) %>%
  round(3) %>%
  kable(caption = "Pearson Correlation Matrix")
```

---

## Planned Analysis

- **Multiple linear regression:** `win_pct ~ net_rtg + ts_pct + pace`
- Interpret r2 to evaluate model fit
- Assess individual predictor significance with p values
- Check regression assumptions such as residual plots, normality, homoscedasticity


